---
title: "Explore Vaccine Trends"
author: "Meg Hutch"
date: "6/11/2021"
output: html_document
---

```{r}
library(tidyverse)
library(ggplot2)
```

```{r Import Vaccine Data}
vaccine_demo_df <- read.csv("data/demographic_trends_of_people_receiving_covid19_vaccinations_in_the_united_states_20210611_clean.csv")

census_data <- read.csv('data/sc-est2020-18+pop-res.csv')
```

**Clean US Census Data**

Calculate new variable ```POPEST17uNDER2020``` for estimated population size of ages $\leq$ 17.

```{r Format Census Data}
census_data <- census_data %>% 
  mutate(NAME = tolower(NAME)) %>%
  rename('State' = NAME) %>% 
  mutate(POPEST17Under2020 = POPESTIMATE2020 - POPEST18PLUS2020) %>%
  rename('Region' = 'REGION')
```

```{r Format Date}
colnames(vaccine_demo_df)[1] <- "Date"
vaccine_demo_df$Date <- as.Date(vaccine_demo_df$Date, format="%m/%d/%Y")
```

```{r Select Age Columns}
vaccine_demo_df <- vaccine_demo_df %>% 
  filter(grepl("Ages", Demographic.Group))
```

For now, we will drop ages < 12 as of June 23, this has not been approved and I'm unsure where these numbers might be coming from (perhaps Clinical Trials but we can not be sure these are successful yet) 

```{r}
vaccine_demo_df <- vaccine_demo_df %>% 
  filter(!Demographic.Group	== "Ages_<12yrs")
```

Here we create new columns:

* ```sum_pediatric```: Sum of all pediatric columns for each day (each row)
* ```sum_adult```: Sum of all pediatric columns for each day (each row)
* ```new_pediatric```: Number of new pediatric vaccines from the previous day
* ```new_adult```: Number of new adult vaccines from the previous day

```{r Pivot Vaccine-Age Table}
vaccine_age_table <- vaccine_demo_df %>% 
  arrange(Date) %>% 
  select(Date, People.who.are.fully.vaccinated, Demographic.Group) %>% 
  pivot_wider(id_cols = Date, 
              names_from = Demographic.Group, 
              values_from = People.who.are.fully.vaccinated) %>% 
  mutate(sum_pediatric = rowSums(.[8:9]), 
         sum_adult = rowSums(.[2:7]),
         new_pediatric = sum_pediatric - lag(sum_pediatric),
         new_adult = sum_adult - lag(sum_adult))
```

Create new variable ```week_id``` which groups dates into 7 day bins starting on Friday-Thursday as the HHS date bins do

```{r Bin dates to weeks}
vaccine_age_table$week_id <- 1+ as.numeric(vaccine_age_table$Date  - as.Date("2020-12-18")) %/% 7
```

Next we group_by week_id and create the following new columns:

* ```new_pediatric_week```: sum of the new pediatric vaccines for that week
* ```new_adult_week```: sum of the new adult vaccines for that week
* ```total_pediatric_week```: the cumulative number of pediatric vaccines by the end of the week
* ```total_adult_week```: the cumualtive number of adult vaccines by the end of the week

```{r Sum counts by week_id}
vaccine_age_table <- vaccine_age_table %>% 
  group_by(week_id) %>% 
  mutate(new_pediatric_week = sum(new_pediatric),
         new_adult_week = sum(new_adult),
         total_pediatric_week = last(sum_pediatric),
         total_adult_week = last(sum_adult)) %>% 
  ungroup()
```

Standardize counts/rates by US Census

```{r Standarize by US Census}
pop_adult_nation <- census_data %>% 
  filter(State == "united states") %>% 
  select(POPEST18PLUS2020) %>% 
  as.numeric()

pop_pediatric_nation <- census_data %>% 
  filter(State == "united states") %>% 
  select(POPEST17Under2020) %>% 
  as.numeric()

vaccine_age_table <- vaccine_age_table %>% 
  mutate(new_pediatric_week_stan = new_pediatric_week/pop_pediatric_nation*100000,
         total_pediatric_week_stan = total_pediatric_week/pop_pediatric_nation*100000,
         new_adult_week_stan = new_adult_week/pop_adult_nation*100000,
         total_adult_week_stan = total_adult_week/pop_adult_nation*100000)
```

Add collection week to harmonize format with HHS data

```{r}
vaccine_age_table <- vaccine_age_table %>% 
  group_by(week_id) %>% 
  mutate(collection_week = first(Date)) %>% 
  filter(!collection_week == "2020-12-16") %>% 
  ungroup()
```

Save processed vaccine and age data

```{r}
save(vaccine_age_table, file = "processed_data/vaccine_age_table.rda")
```


```{r Plot Total Vaccination Rates}
ggplot(vaccine_age_table %>% 
         distinct(week_id, total_pediatric_week_stan, total_adult_week_stan) %>% 
         mutate(total_pediatric_week_stan = total_pediatric_week_stan*5) %>% 
         pivot_longer(!week_id, names_to = "population", values_to = "counts"))  +
       aes(x = week_id, y = counts, group = population, color = population) + 
  geom_line() + 
  geom_point() + 
  ylab("Vaccinations per 100K Adults") + 
  xlab("") + 
  scale_color_brewer(palette="Dark2", labels = c("Adult", "Child"), name = "") + 
  scale_y_continuous(sec.axis = sec_axis(~ ./5, name = "Vaccinations per 100k Children")) + 
  theme_bw() + 
  theme(plot.title = element_text(""),
        legend.title = element_text(face = "bold", size = 15),
        legend.text = element_text(size=15),
        legend.position = "top",
        legend.direction = "horizontal",
        axis.text.x = element_text(face = "bold", size = 10),
        axis.title.y = element_text(face = "bold", size = 13, color = "#1B9E77", margin=margin(0,10,0,0)),
        axis.text.y.left = element_text(face = "bold",  size = 13, color = "#1B9E77"),
        axis.title.y.right = element_text(face = "bold", size = 13, color='#D95F02', margin=margin(0,0,0,10)),
        axis.text.y.right = element_text(face = "bold", size = 13, color='#D95F02')) + 
  theme(strip.text.x = element_text(size = 18))  
```









